High geomagnetic field intensity recorded by anorthosite xenoliths requires a vigorous late Mesoproterozoic geodynamo

### 2021-11-06 PNAS_manuscript/PNAS_manuscript.tex
Updates from Overleaf

\documentclass[draft]{agujournal2019}
\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\usepackage{adjustbox}
\linenumbers

\draftfalse

%% Enter journal name below.
%% Choose from this list of Journals:
%
% JGR: Atmospheres
% JGR: Biogeosciences
% JGR: Earth Surface
% JGR: Oceans
% JGR: Planets
% JGR: Solid Earth
% JGR: Space Physics
% Global Biogeochemical Cycles
% Geophysical Research Letters
% Paleoceanography and Paleoclimatology
% Radio Science
% Reviews of Geophysics
% Tectonics
% Space Weather
% Water Resources Research
% Geochemistry, Geophysics, Geosystems
% Journal of Advances in Modeling Earth Systems (JAMES)
% Earth's Future
% Earth and Space Science
% Geohealth
%

\journalname{JGR: Solid Earth}

@@

\title{High geomagnetic field intensity in the late Mesoproterozoic recorded by Midcontinent Rift anorthosite xenoliths}

\authors{Yiming Zhang\affil{1}, Nicholas L. Swanson-Hysell\affil{1}, Margaret S. Avery\affil{1,3}, Roger R. Fu\affil{4}}

\affiliation{1}{Department of Earth and Planetary Science, University of California, Berkeley, CA, USA}
\affiliation{2}{Department of Earth and Environmental Sciences, University of Minnesota, Duluth, MN, USA}
\affiliation{3}{Geology, Minerals, Energy, and Geophysics Science Center, U.S. Geological Survey, Moffett Field, CA, USA}
\affiliation{4}{Department of Earth and Planetary Sciences, Harvard University, Cambridge, MA, USA}
\correspondingauthor{Yiming Zhang}{yimingzhang@berkeley.edu}

\begin{keypoints}
\item New \textit{ca.} 1092 Ma Midcontinent Rift anorthosite record high paleointensity values
\item High VADM of the geomagnetic field is inconsistent with previous interpretation of Proterozoic long term decay
\item A strong geodynamo like today likely persisted for about 25 Myr during the Midcontinent Rift development
\end{keypoints}

% keywords: absolute paleointensity, Laurentia, Midcontinent Rift, anorthosite, geodynamo

\begin{abstract}
New high-quality absolute paleointensity data from \textit{ca.} 1092 Ma anorthosite xenoliths within the Beaver River diabase intrusions of the North American Midcontinent Rift record an average virtual axial dipole moment (VADM) estimate of $\sim$80 ZAm$^2$. This cooling rate corrected paleointensity estimate from Midcontinent Rift intrusions is similar to previous results from extrusive units that are dated to be between \textit{ca.} 1108 Ma and \textit{ca.} 1085 Ma. Rock magnetic analyses and magnetic imaging reveal that the anorthosite xenoliths can often have magnetic mineralogy dominated by grains with high coercivities, which can yield high success rates passing our selection criteria relative to their diabase hosts. Taken together, new and previously published paleointensity data from the Midcontinent Rift support that a strong geomagnetic field likely persisted through 25 million years of the late Mesoproterozoic. These data are inconsistent with interpretations of a progressive decay of Earth's geodynamo strength through the Proterozoic Eon.

\end{abstract}

\section*{Introduction}

Earth's magnetic field is the result of convective flow of liquid iron-alloy in Earth's outer core. This self-sustaining geodynamo is collectively powered by the heat flow across the core-mantle boundary (CMB) and from the crystallization of the solid inner core from the liquid outer core which provides latent heat and compositional buoyancy. Paleomagnetic studies have found that a dynamo may have existed by the Archean \cite{Selkin2007a, Tarduno2014a, Brenner2020a}. However, some estimates of the thermal conductivity of the core suggest that the nucleation of the inner core could be younger than 1 Ga, more recent than previous thought \cite{Pozzo2012a, Gomi2013a, Labrosse2015a, Ohta2016a}. Without the power of compositional convection provided by the process of light elements being rejected into the outer core from the crystallizing inner core, the geodynamo was likely dominantly driven by thermal conduction through the CMB, with perhaps additional stirring by other power sources such Mg precipitation \cite{ORourcke2016a}, or perhaps there were intervals early in Earth history when the dynamo was generated by exotic processes such as a convecting basal magma ocean \cite{Ziegler2013a}. Long-term secular cooling of Earth has the potential to have decreased the power available to the geodynamo through time prior to the nucleation of the inner core \cite<e.g>{Davies2021a}. As a result, it is possible that rigor of the geomagnetic axial dipole field would have experienced a long-term decay before the emergence of the solid inner core after which it would have strengthened \cite{Aubert2009a, Davies2021a}. However, the estimates for thermal conductivity values for the core continue to vary\cite{Konopkova2016a, Ohta2016a}. Obtaining an observational record of the timing of inner core formation would constrain the thermal evolution of the core and mantle, in addition to improving our understanding of the history of Earth’s dynamo.

@@

%Retrieving paleomagnetic information from Precambrian rocks can aid in our understanding of the long-term evolution of the geodynamo. A recent paleointensity study by \cite{Sprain2018a} reported a \textit{ca.} 1.1 Ga paleomagnetic field strength similar to today based on the volcanics of the Midcontinent Rift System (MRS). That study suggested that this result is consistent with models wherein this time period of the Proterozoic is characterized by a strong rather than weak geomagnetic field. However, to further evaluating the evolution of the geomagnetic field during the Precambrian, more paleointensity data are needed.

%The MRS emplaced a vast amount of extrusive and intrusive rocks spanning a prolonged period of $\sim$25 Myrs. \cite{Swanson-Hysell2019a} published an extensive compilation of the MRS volcanics with paired high-precision CA-ID-TIMS U-Pb geochronology and high-quality paleomagnetic directions, revealing at least two dipole reversals and rapid differential plate motion of the ancient North American plate, Laurentia, during the emplacement of MRS. The synthesized North American apparent polar wander path (APWP) was interpreted to be composed primarily of rapid plate motion \citep{Swanson-Hysell2019a}. Through comparison between paleomagnetic pole positions and the APWP, such rapid plate motion and high-precision paleomagnetic and geochronological framework revealed by the MRS volcanics can provide insights into age constraints on units with poor or no geochronology data. This framework allows for further paleomagnetic contributions beyond the Midcontinent Rift volcanics.

\section*{Geologic Settings}
% a map of the lake superior including the pesonen units, the lake shore traps, the BRD, and the osler and mamainse, together with date compilation.

The Mesoproterozoic North American Midcontinent Rift (MCR) is an intracontinental rift that did not result in the break-up of Laurentia. Although the protracted magmatic activity in the rift lasted form \textit{ca.} 1109 Ma to \textit{ca.} 1084 Ma \cite{Swanson-Hysell2019a}, its activity during the main magmatic stage ($\sim$1098–1090 Ma; \citeA{Vervoort2007a, Miller2013a}) was punctuated by rapid and voluminous emplacement of intrusive rocks as well as extrusive volcanics. Based on high-precision $^{206}$Pb/$^{238}$U geochronology and paleomagnetic data, \citeA{Swanson-Hysell2021a} hypothesized that the massive Duluth Complex and the coeval North Shore Volcanic group (Figure \ref{fig:Geologic_map}) formed as a result of having upwelling mantled-derived plume material encountering Laurentia's thinned lithosphere, following an ``upside-down” drainage" which rapidly funneled melt to today's North Eastern Minnesota (Figure \ref{fig:Geologic_map}; \citeA{Sleep1997a, Swanson-Hysell2014a}). The plagioclase-phyric Anorthositic Series rocks, the crystal-poor Layered Series intrusions of the Duluth Complex, and the coeval volcanics of the North Shore Volcanic Group (NSVG) that outcrop along Northeastern Minnesota (Figure \ref{fig:Geologic_map}) are interpreted to be a result of such vigorous magmatic activity.

\begin{figure}
\centering
\noindent\includegraphics[width=\textwidth]{Geologic_map.pdf}
\caption{\footnotesize{Simplified geologic map of Lake Superior region showing the distribution of rift and post-rift-related rocks including extrusive, intrusive, and late/post-rift sediments around Lake Superior. Red diamonds mark sites with paleointensity results that passed the selection criteria from this study. Paleomagnetic sites from \citeA{Pesonen1983a} are categorized by lithology. All sites from previous studies are color-coded by their ages.}}
\label{fig:Geologic_map}
\end{figure}

The emplacement of the \textit{ca.} 1092 Ma Beaver Bay Complex punctuates another period of magmatism during the Midcontinent Rift main stage activity. $^{206}$Pb/$^{238}$U zircon dates and cross-cutting relationships tightly constraints the areally extensive Beaver River diabase of the Beaver Bay Complex to have been emplaced at 1091.7 $\pm$ 0.2 Ma \cite{Zhang2021b}. The Beaver River diabase intruded into hypabyssal depths via wide conduits that accommodated the entrainment of numerous anorthosite xenoliths that have short-axis diameters up to 180 meters \cite{Boerboom2004a, Boerboom2006b}. Geochronological, paleomagnetic, geochemical, and petrographic data are consistent with the hypothesis that the Beaver River diabase magma extruded to the surface, feeding the massive Greenstone Flow - one of the largest lava flows on Earth \cite{Doyle2016a, Zhang2021b}.

% The anorthosite xenoliths that distinctly occur within the Beaver River diabase are not only recorders of the wide diabase conduit sizes,  of the geomagnetic field $\sim$1091.7 million years ago. Thermal modeling result and paleomagnetic directional data show the anorthosite xenoliths acquired thermal remanent magnetizations during cooling of the diabase \cite{Zhang2021b}. Step-wise thermal demagnetizations show many anorthosite xenoliths have minimal secondary remanence and single-component characteristic remanence magnetizations that often unblock sharply within temperature ranges between 500\textdegree C and 580\textdegree C.

Despite that the anorthosite xenoliths of the Beaver River diabase were entrained as xenoliths and emplaced only $\sim$4 Myr after the formation of the Anorthositic Series rocks of the Duluth Complex when massive plagioclase mushes were generated, $^{206}$Pb/$^{238}$U zircon geochronology data and zircon rare earth element diffusion modeling results suggest it is the most probable that the Beaver River diabase anorthosite xenoliths are entrained cumulate enclaves that formed at the time of Beaver Bay Complex magmatism. Moreover, the anorthosite xenoliths of the \textit{ca.} 1092 Ma Beaver River diabase has distinct mineralogy from that of the Anorthositic Series rocks of the Duluth Complex. The dominantly monomineralic anorthosite xenoliths have an average anorthite content of $\sim$70\% \cite{Morrison1983a, Doyle2016a}, higher than the average of $\sim$60\% of the Anorthositic Series rocks of the Duluth Complex \cite{Miller1990a}. In addition, petrographic imaging support that the plagioclase from the Beaver River anorthosite formed under a magmatic conditions different from the Duluth Complex anorthositic series magma. While the plagioclase in the Duluth Complex Anorthositic Series typically are euhedral with an interlocking texture (Figure \ref{Petro_QDM}A), the more calcium-rich plagioclase of the Beaver River anorthosite often develop granoblastic texture characterized by equigranular plagioclase crystals (Figure \ref{Petro_QDM}B). Abundant magnetite and ilmenite needles that are typically 10 $\mu$m long can be found to have preferred orientations within Duluth Complex plagioclase. Interstitial vermicular titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex Anorthositic Series rocks \cite{Miller1990a}.

\begin{figure}
\noindent\includegraphics[width=\textwidth]{Petro_QDM.png}
\caption{\footnotesize{Thin section petrographic and magnetic images of anorthosites from Duluth Complex Anorthositic Series (left column) and Beaver River anorthosite xenolith from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (right column; \citeA{Zhang2021b}). Petrographic images (A-D) show Duluth Complex anorthosite have euhedral, elongate plagioclase crystals with interlocking texture and igneous foliation. (C) The plagioclase can contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the (010) planes of plagioclase. In contrast, plagioclase of sample MS99033 do not enclose large oxide needles. Magnetic field maps acquired by the quantum diamond microscope (QDM) show that the oxide needles within Duluth Complex plagioclase are strongly magnetic as evidenced through the overlapping magnetic dipole fields exerted above the oxides in (E). In comparison, relatively weak magnetic sources within and interstitial to plagioclase crystals exist in the Beaver River anorthosite. The dipole-like features in (F) do not show strong preferred orientations. Magnetic field maps of (E) and (F) share the same spatial and magnetic scale. The unit of the axes are in $\mu$m.}}
\label{fig:Petro_QDM}
\end{figure}

Occurrences of magnetite and ilmenite symplectites and needles within anorthositic rocks have been described by \citeA{Scofield1975a, Scofield1986a} in the Wichita Mountains and by \citeA{Haggerty1976a} in other slowly cooled, mafic intrusions. \citeA{Scofield1986a} hypothesized that the inclusions acted as a means of buffering the melt with respect to oxygen, forming either crystallites along preexisting plagioclase substrate or as iron incorporated into the plagioclase that subsequently exsolved at subsolidus temperatures. Such a model requires that plagioclase, instead of other mafic minerals such as olivine, to be the first to crystallize from a magma and act as a sink for iron. The two-stage formation of the Duluth Complex Anorthositic Series fits this model \cite{Miller1990a}. During the first stage, high pressure at subcrustal depths (15-25 km depth) caused fractionation of Duluth Complex magma resulting in the formation of floating plagioclase mushes \cite{Kushiro1980a}. The second stage of crystallization then occurs at low pressures after the magma intruded into shallower chambers.

This proposed magmatic evolution of the Duluth Complex is in distinct contrast to the emplacement of the Beaver River diabase which rapidly ascended and entrained the anorthosite xenoliths directly into shallow chambers \cite{Flower1980a, Miller1997a, Zhang2021b}. Particularly, \citeA{Miller1997a} observed a high state of structural disorder within plagioclase of the Beaver River anorthosite using X-ray diffraction analyses. This is consistent with the interpretation that these anorthosite were rapidly transported by upward currents into hypabyssal magma chambers without reaching structural equilibrium. Within the Beaver River anorthosite xenoliths, the titanomagnetite needles and the symplectic magnetite-ilmenite intergrowths are absent (Figure \ref{Petro_QDM}). Instead, relatively sparse titano-magnetite within or interstitial to the plagioclase are typical of Beaver River anorthosites (Figure \ref{Petro_QDM}).

% Previous studies have divided magmatic activity in the rift into four stages based on interpreted changes in relative magmatic volume and the nature of magmatism: early ($\sim$1109–1104 Ma), latent ($\sim$1104–1098 Ma), main ($\sim$1098–1090 Ma) and late ($\sim$1090–1084 Ma)  In particular, the main stage magmatism is characterized by rapid and punctuated events which resulted in voluminous emplacement of intrusive rocks as well as coeval outpouring of extrusive lava flows. One example is the massive \textit{ca.} 1096 Ma Duluth Complex and the North Shore Volcanic Group \cite{Swanson-Hysell2020a}, the majority of which emplaced within $<$500 kyrs. Another one is the \textit{ca.} 1092 Ma Beaver Bay Complex. The Beaver River diabase intrusions of the complex formed synchronously with the Greenstone Flow of the Portage Lake Volcanics which is one of the largest lava flows known on Earth \cite{Zhang2021b}. Excellent rock preservation thanks to minimal post-rift alteration in the Midcontinent Rift have yielded high-quality paleomagnetic directional data which have now been paired with high-precision zircon geochronology to reconstruct the ``Keweenawan Track" - a well-resolved Laurentia's (cratonic North America) apparent polar wander path during the late Mesoproterozoic \cite{Swanson-Hysell2019a}.

\section*{Method}

\subsection*{Sample collection and paleomagnetic directions}

To further identify the magnetic carriers within the Beaver River anorthosite xenoliths and compare them with the anorthosites of the Duluth Complex Anorthositic Series rocks, we used the quantum diamond microscope (QDM) at the UC Berkeley Paleomagnetism laboratory to image a thin section of sample MS99033 which is a hand sample from anorthosite xenolith AX16 (which yielded a $^{206}$Pb/$^{238}$U zircon date of 1091.83 $\pm$ 0.21 Ma; \citeA{Zhang2021b}), and a thin section of a Duluth Complex anorthosite (Figure \ref{fig:Petro_QDM}). We the QDM to image the magnetic field over the polished thin section surfaces at a sample-sensor distance of 5 $\mu$m in projective magnetic microscopy (PMM) mode with a spatial resolution of 4.7 μm per pixel and an instantaneous 0.9 mT bias field that is canceled during the course of measurement \cite{Glenn2017a}).

\section*{Results and Discussion}
\subsection*{Paleointensity}

@@

@@

\subsection*{Late Mesoproterozoic geodynamo }
To best characterize the geomagnetic axial dipole field intensity during a certain time period requires that a paleointensity dataset to cover a sufficient amount of time such that paleosecular variations of the geomagnetic field are averaged and the geocentric axial dipole hypothesis (GAD) can be assumed. Thermal modeling results from \citeA{Zhang2021b} suggest that the anorthosite xenoliths were likely heated to tholeiitic magma temperature and acquired thermal remanent magnetization during cooling with their diabase host on a time scale of a few thousand years. In addition, the closely located sites AX12 and AX16 have very similar paleointensity estimates that are distinct from those of sites AX11 and AX13 which were emplaced $\sim$10 km away (Figure \ref{fig:Geologic_map}). The consistency within small regions but variations between regions could be the result of differential cooling history due to heterogeneity of the timing and depths of the Beaver River diabase intrusions, causing the anorthosites to capture paleosecular variations \textit{ca.} 1092 Ma. Overall, none of the anorthosite xenoliths record a V(A)DM value smaller than 40 ZAm$^2$ (Figure \ref{fig:PINT_compilation}).

@@

% 1. whether septils was recording the time averaged field or the low values are from the hyperactive state which is another signal prior to the onset of inner core? indeed it is interesting that no high values from the Ediacaran has been recovered

% 2. where is the recovery, how long was the recovery, how strong was the recovery is at question for he Ediacaran boundary.

% 3. How should one explain the Mundine Well dikes high values?

% 4. How should one explain the Bungemall sills low values in 1070 Ma following the high values of the MCR?

% 5. We need more spatial and temporal coverage of the Precambrian observational data.

% using paleomagnetic observational records is guided by the predictions from numerical simulations that a cooling liquid core with only thermal convection power to drive a dynamo would experience a progressive decay in its magnetic dipole field strength, and experience a rapid recover after the onset of inner core nucleation \cite{Davies2021a}. Previous studies has pointed to the Ediacaran-Cambrian boundary for such a weakening field behavior and recent studies has claim to found a rapid increase after the single crystal time which fits in the numerical models. However, such a recover may not be unique given the strong geodynamo estimate from the 1.1 Ga Midcontinent Rift. One could point to the increase from low values from the ca. 1140 Ma abitibi dikes to the ca. 1092 Ma anorthosites to be the result of recovery; another line of disagreement is between the Bangemall sills and the low field from Sept Ils. MCR being an anomaly? or the time scale we are looking at is too long?

% use a combined method of direction and intensity research for further examinations. Re-evaluation of the spatial and temporal variability captured by the existing Precambrian database. More data are needed for the Precambrian in order to better characterize the geodynamo evolution.

% question to ask about the paleointensity database:

@@

% cooling rate correction problem - that the Nagy et al paper can suggest even less of a correction and higher of the estimated field. theoretical thermal modeling by \citeA{Zhang2021a} suggest tha the anorthosite xenoliths were likely heated to diabase magma temperature regardless of their sizes, and cooled through the typical characteristic unblocking temperature range from 580 to 500 on the order of 1000 years. Comparing to the typical lab cooling time of about 100 seconds duirng paleointensity experiments, the significantly slower cooling rate when the NRM was acquired as the anorthosite xenoliths cooled with the diabase led to overestimate of paleointensity by up to 40\%. Overall range of correction factor can be 1.1-1.4, in any case consistent with a high field strength

% that \citeA{bono2019a} proposed an interesting trend but has problems in the directions, likely to be a short term feature of a low field during a reversal, instead of a low field as a result of long term cooling.

% that Lloyd 2021 also presented low field that fits the trend, but they similarly have the issue of reversal, and that they do not have good control on secular variations

% Together with previously published data, the Midcontinent Rift rocks are recording 1.1 Ga earth geomagnetic field that is stronger than the mean of the past 300 Ma (Figure \ref{}).

% Regardless, a strong geomagnetic field has been recorded by the \textit{ca.} 1092 Ma Beaver River anorthosite xenoliths and a strong geodynamo might have persisted through the protracted Midcontinent Rift magmatism.

@@

\acknowledgments
Project research was supported by NSF CAREER grant EAR-1847277 to N.L.S.-H. and an Institute on Lake Superior Geology Student Research Fund grant to Y.Z. Permits for fieldwork and sampling from the Minnesota Department of Natural Resources are gratefully acknowledged. We thank James Pierce and Blake Hodgin for their assistance in the field. We thank Jim Miller for providing us with the Duluth Complex anorthosite thin section. We thank Dario Bilardello, Peat Solheid, Mike Jackson, Josh Feinberg, and Bruce Moskowitz for their tremendous help of instrumental operations, data interpretations, and research guidance. Y.Z. thanks the IRM for the U.S. Visiting Student Fellowship. Paleomagnetic data associated with this study are available within the MagIC database (\url{}; UPDATE WHEN DOI IS GENERATED) and all data are within a Github repository associated with this work (\url{}) that is also archived on Zenodo (INSERT URL AT TIME OF PROOFS). This repository also contains Python code related to calculations, visualizations and statistical tests discussed herein.  
\bibliography{YZ_ref}

@@

% potential reviewer: Peter Selkin, Jeff Gee, Lisa Tauxe, Roger Fu, Lauri Pesonen, Josh Feinberg Julie Bowles,

% At the IRM, I was seeking the answer to this question with the help of the vibrating sample magnetometer (VSM) systems, including a newly installed Lake Shore VSM which greatly helped improve measurement resolution on samples with weak magnetizations (Figure \ref{fig:backfield}). Backfield demagnetization experiments were conducted and used to develop coercivity spectra (Figure \ref{fig:backfield}). The spectra were subsequently modeled to fit for the distributions of different  populations of magnetic particles using similar procedure as \cite{Maxbauer2016a}. Most spectra can be well approximated with models with one component or two components with overlapping coercivity ranges (Figure \ref{fig:backfield}). The dominance of single-component distributions from the unmixed coercivity spectra likely suggests minimal alterations or formation of secondary magnetic mineral within the samples. By compiling the values of median destructive field (MDF) from all measured specimens and categorizing them in terms of their sister specimens' paleointensity results, I found that the anorthosites samples that pass paleointensity selection criteria have distinctively higher MDF values than those did not (Figure \ref{fig:backfield}), and all diabase specimens have low MDF values like the non-ideal anorthosites (Figure \ref{fig:backfield}). The higher MDFs ($\sim$60 mT) of the anorthosites are similar to what is typical of stoichiometric, single-domain magnetite grains which are favored for paleointensity experiments. Furthermore, the low MDFs displayed in other specimens could be linked to a dominant population of large interstitial Fe-Ti oxides that are more likely to form multi-domain grains, which have been observed in both the anorthosite and diabase (Figure \ref{fig:field}). In addition, preliminary tests on single anorthosite crystals might suggest that not all single crystals are necessarily better for paleointensity research when comparing to bulk samples, given their similar low MDF to the failed diabase and anorthosite specimens. However, more single crystal grains with paired paleointensity data and rock magnetic data may be needed before a general conclusion can be drawn.

>>>>

\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
% Use the lineno option to display guide line numbers if required.

\templatetype{pnasresearcharticle} 
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} %= Template for a one-column mathematics article
% {pnasinvited} %= Template for a PNAS invited submission

\title{High geomagnetic field intensity recorded by anorthosite xenoliths requires a vigorous late Mesoproterozoic geodynamo}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{Yiming Zhang}
\author[a,1]{Nicholas L. Swanson-Hysell} 
\author[a,b]{Margaret S. Avery}
\author[c]{Roger R. Fu}

\affil[a]{Department of Earth and Planetary Science, University of California, Berkeley, CA, 94720}
\affil[b]{Geology, Minerals, Energy, and Geophysics Science Center, U.S. Geological Survey, Moffett Field, CA, 94025}
\affil[c]{Department of Earth and Planetary Sciences, Harvard University, Cambridge, MA, 02138}

% Please give the surname of the lead author for the running footer
\leadauthor{Zhang}

% Please add a significance statement to explain the relevance of your work
\significancestatement{Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality. The primary goal of the significance statement is to explain the relevance of the work in broad context to a broad readership. The significance statement appears in the paper itself and is required for all research papers.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{Author contributions: Y.Z. and N.L.S.-H. designed research; Y.Z., N.L.S.-H. and M.S.A studied and sampled the rock cores; Y.Z. and N.L.S.-H. conducted and analyzed rock magnetic experiments; Y.Z. performed petrography and electron microscopy; R.R.F. contributed to magnetic imaging data analyses; and Y.Z., N.L.S.-H., M.S.A, and R.R.F. wrote the paper.}

\authordeclaration{The authors declare no conflict of interest.}
% \equalauthors{\textsuperscript{1}A.O.(Author One) contributed equally to this work with A.T. (Author Two) (remove if not applicable).}
\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: yimingzhang\@berkeley.edu}

% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
\keywords{absolute paleointensity $|$ Laurentia $|$ Midcontinent Rift $|$ anorthosite $|$ geodynamo $|$}

\begin{abstract}
Obtaining estimates of Earth's magnetic field strength in deep time is complicated by non-ideal rock magnetic behavior in many igneous rocks. In this study, we target \textit{ca.} 1092 Ma anorthosite xenoliths that cooled and acquired their magnetization within the Beaver River diabase intrusions of the North American Midcontinent Rift. In contrast to the diabase which

These anorthosite xenoliths cooled and acquired their magnetization at the same time as their diabase hosts, but their magnetic

New high-quality absolute paleointensity data from  record virtual axial dipole moments (VADM) estimates of $\sim$40 to 130 ZAm$^2$. These cooling rate corrected paleointensity estimates are similar to and higher than previous results from extrusive units that are dated to be between \textit{ca.} 1108 Ma and \textit{ca.} 1085 Ma. Rock magnetic analyses and magnetic imaging reveal that the anorthosite xenoliths can often have magnetic mineralogy dominated by grains with high coercivities, which can yield high success rates passing our selection criteria relative to their diabase hosts. Taken together, new and previously published paleointensity data from the Midcontinent Rift support that a strong geomagnetic field likely persisted through 25 million years of the late Mesoproterozoic. These data are inconsistent with interpretations of a continuous progressive decay of Earth's geodynamo strength through the Proterozoic Eon. 
\end{abstract}

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

@@

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{E}arth's magnetic field is the result of convective flow of liquid iron-alloy in Earth's outer core. At present day, the geodynamo is collectively powered by heat flow across the core-mantle boundary (CMB) and from the crystallization of the solid inner core from the liquid outer core which provides latent heat and compositional buoyancy. However,  while paleomagnetic studies have found that a dynamo field has existed since at least 3.4 billion years ago in the Archean \cite{Selkin2007a, Tarduno2014a, Brenner2020a}, Earth's inner core likely crystallized more recently. Without an inner core, Earth's magnetic field would have been generated solely through CMB heat flow. Estimates of the timing of the initial crystallization of Earth's inner core are interconnected with estimates for the core's thermal conductivity. Higher conductivity values imply faster cooling rates which in turn imply that the threshold for the freezing of the inner core happened more recently \cite{Davies2015a}. While estimates of these values were once consistent with a ca. 3.5 Ga age for the inner core \cite{Gubbins2004a}, revised estimates based on first-principles calculations implied higher conductivities \cite{Koker2012a,Pozzo2012a} and an age for the inner core that is less than 1 Ga \cite{Pozzo2012a, Gomi2013a, Labrosse2015a, Ohta2016a}. Without the power of compositional convection provided by the process of light elements being rejected into the outer core from the crystallizing inner core, the geodynamo was likely dominantly driven by thermal conduction through the CMB, with perhaps additional stirring by other power sources such Mg precipitation \cite{ORourke2016a}, or perhaps there were intervals early in Earth history when the dynamo was generated by exotic processes such as a convecting basal magma ocean \cite{Ziegler2013a}. Long-term secular cooling of Earth has the potential to have decreased the power available to the geodynamo through time prior to the nucleation of the inner core \cite<e.g>{Davies2021a}. As a result, it is possible that intensity of the geomagnetic axial dipole field would have experienced a long-term decay before the emergence of the solid inner core after which it would have strengthened \cite{Aubert2009a, Davies2021a}. However, the estimates for thermal conductivity values for the core continue to vary \cite{Konopkova2016a, Ohta2016a}. Obtaining an observational record of the timing of inner core formation would constrain the thermal evolution of the core and mantle, in addition to improving our understanding of the history of Earth’s dynamo.

@@

\section*{Results and Discussion}

\subsection*{Paleointensity}

@@

Many authors find it useful to organize their manuscripts with the following order of sections;  title, author line and affiliations, keywords, abstract, significance statement, introduction, results, discussion, materials and methods, acknowledgments, and references. Other orders and headings are permitted.

@@

\section*{Late Mesoproterozoic geodynamo }
To best characterize the geomagnetic axial dipole field intensity during a certain time period requires that a paleointensity dataset to cover a sufficient amount of time such that paleosecular variations of the geomagnetic field are averaged and the geocentric axial dipole hypothesis (GAD) can be assumed. Thermal modeling results from \citeA{Zhang2021b} suggest that the anorthosite xenoliths were likely heated to tholeiitic magma temperature and acquired thermal remanent magnetization during cooling with their diabase host on a time scale of a few thousand years. In addition, the closely located sites AX12 and AX16 have very similar paleointensity estimates that are distinct from those of sites AX11 and AX13 which were emplaced $\sim$10 km away (Figure \ref{fig:Geologic_map}). The consistency within small regions but variations between regions could be the result of differential cooling history due to heterogeneity of the timing and depths of the Beaver River diabase intrusions, causing the anorthosites to capture paleosecular variations \textit{ca.} 1092 Ma. Overall, none of the anorthosite xenoliths record a V(A)DM value smaller than 40 ZAm$^2$ (Figure \ref{fig:PINT_compilation}).

@@

@@

@@

\matmethods{
\subsection*{Sample collection and paleomagnetic directions}

To further identify the magnetic carriers within the Beaver River anorthosite xenoliths and compare them with the anorthosites of the Duluth Complex Anorthositic Series rocks, we used the quantum diamond microscope (QDM) at the UC Berkeley Paleomagnetism laboratory to image a thin section of sample MS99033 which is a hand sample from anorthosite xenolith AX16 (which yielded a $^{206}$Pb/$^{238}$U zircon date of 1091.83 $\pm$ 0.21 Ma; \citeA{Zhang2021b}), and a thin section of a Duluth Complex anorthosite (Figure \ref{fig:Petro_QDM}). We the QDM to image the magnetic field over the polished thin section surfaces at a sample-sensor distance of 5 $\mu$m in projective magnetic microscopy (PMM) mode with a spatial resolution of 4.7 μm per pixel and an instantaneous 0.9 mT bias field that is canceled during the course of measurement \cite{Glenn2017a}).
}

\showmatmethods{} % Display the Materials and Methods section

\acknow{Project research was supported by NSF CAREER grant EAR-1847277 to N.L.S.-H. and an Institute on Lake Superior Geology Student Research Fund grant to Y.Z. Permits for fieldwork and sampling from the Minnesota Department of Natural Resources are gratefully acknowledged. We thank James Pierce and Blake Hodgin for their assistance in the field. We thank Jim Miller for providing us with the Duluth Complex anorthosite thin section. We thank Dario Bilardello, Peat Solheid, Mike Jackson, Josh Feinberg, and Bruce Moskowitz for their tremendous help of instrumental operations, data interpretations, and research guidance. Y.Z. thanks the IRM for the U.S. Visiting Student Fellowship. Paleomagnetic data associated with this study are available within the MagIC database (\url{}; UPDATE WHEN DOI IS GENERATED) and all data are within a Github repository associated with this work (\url{}) that is also archived on Zenodo (INSERT URL AT TIME OF PROOFS). This repository also contains Python code related to calculations, visualizations and statistical tests discussed herein.}

\showacknow{} % Display the acknowledgments section

% Bibliography
\bibliography{YZ_ref}

@@

% \subsection*{Manuscript Length}

% A standard 6-page article is approximately 4,000 words, 50 references, and 4 medium-size graphical elements (i.e., figures and tables). The preferred length of articles remains at 6 pages, but PNAS will allow articles up to a maximum of 12 pages.

% \subsection*{References}

% References should be cited in numerical order as they appear in text; this will be done automatically via bibtex, e.g. \cite{belkin2002using} and \cite{berard1994embedding,coifman2005geometric}. All references cited in the main text should be included in the main manuscript file.

% \subsection*{Data Archival}

% PNAS must be able to archive the data essential to a published article. Where such archiving is not possible, deposition of data in public databases, such as GenBank, ArrayExpress, Protein Data Bank, Unidata, and others outlined in the \href{https://www.pnas.org/page/authors/journal-policies#xi}{Information for Authors}, is acceptable.

% \begin{SCfigure*}[\sidecaptionrelwidth][t]
% \centering
% \includegraphics[width=11.4cm,height=11.4cm]{frog}
% \caption{This legend would be placed at the side of the figure, rather than below it.}
% \label{fig:side}
% \end{SCfigure*}

% \subsection*{Digital Figures}

% EPS, high-resolution PDF, and PowerPoint are preferred formats for figures that will be used in the main manuscript. Authors may submit PRC or U3D files for 3D images; these must be accompanied by 2D representations in TIFF, EPS, or high-resolution PDF format. Color images must be in RGB (red, green, blue) mode. Include the font files for any text.

% Images must be provided at final size, preferably 1 column width (8.7cm). Figures wider than 1 column should be sized to 11.4cm or 17.8cm wide. Numbers, letters, and symbols should be no smaller than 6 points (2mm) and no larger than 12 points (6mm) after reduction and must be consistent.

% Figures and tables should be labelled and referenced in the standard way using the \verb|\label{}| and \verb|\ref{}| commands.

% Figure \ref{fig:frog} shows an example of how to insert a column-wide figure. To insert a figure wider than one column, please use the \verb|\begin{figure*}...\end{figure*}| environment. Figures wider than one column should be sized to 11.4 cm or 17.8 cm wide. Use \verb|\begin{SCfigure*}...\end{SCfigure*}| for a wide figure with side legends.

% \subsection*{Tables}
% Tables should be included in the main manuscript file and should not be uploaded separately.
